Fusing Filters with Integer Linear Programming

### 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex
Minor fix
transducer parents

\begin{tabbing}
MMMMM       \= MM \= MMMMMMMMMMM \= \kill
$T$          \> @::=@ \> $size$                                  \> (known size) \\
             \> $~|$  \> @external@                              \> (external and unfusible) \\
\end{tabbing}

@@

After the generated equality constraints are solved, and sizes are grouped into equivalence classes, the combinators with iteration sizes in the same equivalence class can be fused together.

@@

Looking back at the @normalize@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$.
Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
\begin{code}
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ...
\end{code}

% \ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
% No,
We now define a function $trans$, to find the parent transducer of a given combinator.

@@

%\begin{lemma}
\textbf{Lemma: unique transducers}.
For some bindings $bs$, each name $n$ will have at most one transducer.
\[
valid(\constr{bs}) \implies \forall n.\ |trans(bs, n)| \le 1
\]
\textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
The other cases are trivial.
%\end{lemma}

>>>>

\begin{tabbing}
MMMMM       \= MM \= MMMMMMMMMMM \= \kill
$T$          \> @::=@ \> $\tau$                                  \> (known size) \\
             \> $~|$  \> @external@                              \> (external and unfusible) \\
\end{tabbing}

@@

After the generated equality constraints are solved, and sizes are grouped into equivalence classes, combinators with iteration sizes in the same equivalence class may be fused together if there are no fusion-preventing dependencies between them.

@@

Looking back at the @normalize@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
%The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
%\gabi{gts is no combinator, so can't be a transducer
The iteration size of @sum2@ is $k_{gts}$, and the filter combinator which produces @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
\begin{code}
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ...
\end{code}

% \ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
% No,
We now define a function $trans$, to find the parent transducer of a given combinator.
The parent transducer $trans(bs, n)$ of a combinator has the same output size as $n$'s iteration size, but the two have different iteration sizes.

@@

To determine whether two combinators of different iteration sizes may be fused together, we first find parent or ancestor transducers of the same size, if they exist:
\begin{tabbing}
MMMM \= M \= MMMMMMM \= M \= \kill
$parents$ \> $::$ \> $name \to name \to \{name \times name\}$ \\
$parents(a,b)$ \\
        \> $|$ \> $\tau(a) == \tau(b)$ \> $=$    \> $\{(a, b)\}$ \\
        \> $|$ \> otherwise            \> $=$    \> $\{ parents(a', b) ~|~ a' \in trans(a) \} $      \\
        \>     \>                      \> $\cup$ \> $\{ parents(a, b') ~|~ b' \in trans(b) \} $  \\
\end{tabbing}

\textbf{Lemma: sole transducers}.
For some bindings $bs$, each name $n$ will have at most one transducer.
\[
valid(\constr{bs}) \implies \forall n.\ |trans(bs, n)| \le 1
\]
\textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
The other cases are trivial.

\textbf{Lemma: transducers change types}.
For some bindings $bs$, if $n$ has a transducer, the iteration size is not the same.
\[
valid(\constr{bs}) \wedge \{m\} = trans(bs, n) \implies \tau(n) \not= \tau(m)
\]

\textbf{Lemma: sole parents}.
For some bindings $bs$, each pair of names $a$ and $b$ will have at most one set of parents $parents(a,b)$.
\[
valid(\constr{bs}) \implies \forall a,b.\ |parents(bs, n)| \le 1
\]
\textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.
